---
title: "Intermediate \\texttt{R} Programming"
author: "Jake S. Truscott, Ph.D"
subtitle: |
  POS6933: Computational Social Science
institute: |
  \vspace{-5mm} University of Florida \newline
  Spring 2026 \newline \newline \newline
  \includegraphics[width=3cm]{../../beamer_style/UF.png} \quad 
  \includegraphics[width=3.1cm]{../../images/CSS_POLS_UF_Logo.png}
output: 
  beamer_presentation:
    slide_level: 3
    keep_tex: true
    toc: false
classoption: aspectratio=169  
header-includes:
  - \PassOptionsToClass{aspectratio=169}{beamer}
  - \usepackage{../../beamer_style/beamer_style}
  - \setbeamersize{text margin left=3.5mm,text margin right=3.5mm}
---

```{r setup, include=FALSE}
library(dplyr); library(ggplot2); library(cowplot); library(formatR); library(stargazer)

knitr::opts_chunk$set(
  warning = FALSE, 
	fig.align = "center",
	comment = NA,
	dev = "pdf",
	size = "tiny",
	tidy = TRUE, 
  tidy.opts=list(width.cutoff=50)
)

default_ggplot_theme  <- theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12),
    axis.title = element_text(size = 12, colour = 'black'),
    axis.text = element_text(size = 10, colour = 'black'),
    panel.background = element_rect(linewidth = 1, colour = 'black', fill = NA), 
    legend.position = 'bottom', 
    legend.background = element_rect(linewidth = 1, colour = 'black', fill = NA), 
  )

```

# Overview

### Overview

-   \textbf{Today's Goal}: Improve Effectiveness w/ \texttt{R} Programming

    \par \vspace{2.5mm}

-   Random Number Generation in \texttt{R}

    \par  \vspace{2.5mm}

-   Loops and Iteration \vspace{2.5mm}

    \par

-   Visualizing Data and Relationships Using \texttt{ggplot::()} \& \texttt{Stargazer}

### Getting Started

- Navigate to main directory folder w/ \texttt{R} environment \& 3 folders (data, code, practice_set) \par \vspace{2.5mm}
- Open the \texttt{R} environment, then \texttt{File} $\rightarrow$ \text{New File + R Script} \par \vspace{2.5mm}
- Run the code emailed earlier today -- this will download code from GitHub walkthroughs

# Random Number Generation

### Coin Flips

\centering

\textbf{What is the probability that any independent coin flip will land on heads?} \pause

\par \vspace{5mm}

\textbf{Does this change if I flip 50 times?} \pause

\par \vspace{5mm}

\textbf{What about 100 times?}\pause

\par \vspace{5mm}

\textbf{What about 1000 times?}\pause

\par \vspace{5mm}

\textbf{What about 10000 times?}

### Coin Flips (Cont.)

\small

```{r coin_10, echo=FALSE, fig.height=3, fig.width=4, tidy=TRUE}
options(width = 55)
set.seed(1234) # Set Random Seed
coin_10 <- data.frame(side = sample(c('Heads', 'Tails'), size = 10, replace = TRUE)) %>%
  ggplot(aes(x = side, group = side)) + 
  geom_bar(aes(fill = side), colour = 'black') + labs(x ='', y = '', title = '10 Flips') + 
  geom_hline(yintercept = 0) + default_ggplot_theme + theme(legend.position = 'none', axis.text.y = element_blank())+ 
  scale_y_continuous(expand = expansion(mult = c(0, 0.25))) + 
  geom_label(stat = "count", aes(label = after_stat(count)), vjust = 1.5)

coin_10

```

### Coin Flips (Cont.)

\small

```{r coin_100, echo=FALSE, fig.height=3, fig.width=4, tidy=TRUE}
options(width = 55)
set.seed(1234) # Set Random Seed
coin_100 <- data.frame(side = sample(c('Heads', 'Tails'), size = 100, replace = TRUE)) %>%
  ggplot(aes(x = side, group = side)) + 
  geom_bar(aes(fill = side), colour = 'black') + labs(x ='', y = '', title = '100 Flips') + 
  geom_hline(yintercept = 0) + default_ggplot_theme + theme(legend.position = 'none', axis.text.y = element_blank())+ 
  scale_y_continuous(expand = expansion(mult = c(0, 0.25))) + 
  geom_label(stat = "count", aes(label = after_stat(count)), vjust = 1.5)

coin_100

```

### Coin Flips (Cont.)

\small

```{r coin_1k, echo=FALSE, fig.height=3, fig.width=4, tidy=TRUE}
options(width = 55)
set.seed(1234) # Set Random Seed

coin_1k <- data.frame(side = sample(c('Heads', 'Tails'), size = 1000, replace = TRUE)) %>%
  ggplot(aes(x = side, group = side)) + 
  geom_bar(aes(fill = side), colour = 'black') + labs(x ='', y = '', title = '1k Flips') + 
  geom_hline(yintercept = 0) + default_ggplot_theme + theme(legend.position = 'none', axis.text.y = element_blank())+ 
  scale_y_continuous(expand = expansion(mult = c(0, 0.25))) + 
  geom_label(stat = "count", aes(label = after_stat(count)), vjust = 1.5)


coin_1k

```

### Coin Flips (Cont.)

\small

```{r coin_10k, echo=FALSE, fig.height=3, fig.width=4, tidy=TRUE}
options(width = 55)
set.seed(1234) # Set Random Seed
coin_10k <- data.frame(side = sample(c('Heads', 'Tails'), size = 10000, replace = TRUE)) %>%
  ggplot(aes(x = side, group = side)) + 
  geom_bar(aes(fill = side), colour = 'black') + labs(x ='', y = '', title = '10k Flips') + 
  geom_hline(yintercept = 0) + default_ggplot_theme + theme(legend.position = 'none', axis.text.y = element_blank()) + 
  scale_y_continuous(expand = expansion(mult = c(0, 0.25))) + 
  geom_label(stat = "count", aes(label = after_stat(count)), vjust = 1.5)

coin_10k

```

### Coin Flips (Cont.)

-   We can use \texttt{sample()} to randomly select elements from a vector
-   In this case, a coin flip where $p(heads) = p(tails) = 0.5$ \small

```{r uniform_distribution, tidy=TRUE}
sides <- c('Heads', 'Tails') # Flip Options
single_flip <- sample(sides, size = 1) # Single Draw
print(single_flip) 
```

### 6-Sided Die

-   We can use the same approach to "roll" a six-sided die. \small

```{r six_sided_die, tidy=TRUE}

sides <- c(1:6) # 1, 2, 3, 4, 5, 6
single_roll <- sample(sides, size = 1) # Single Roll
message('Result of Single Roll: ', single_roll)

```

\small

### Poker Hands {.fragile}

-   We can even use it to do more complex operations like simulate a random draw from 5-card Poker \tiny

```{r poker, tidy=TRUE}

cards <- as.character(c(2:10, 'J', 'Q', 'K', 'A')) 
# All Card Values
suits <- c('Hearts', 'Diamonds', 'Spades', 'Clubs') 
# Suits

deck <- expand.grid(value = cards, 
                               suit = suits) %>%
  mutate(card = paste(value, "of", suit)) %>%
  pull(card) # Create a Full Deck

random_draw <- sample(deck, size = 5, replace = F) 
# Random 5-Card Draw w/out Replacement
```

### Poker Hands (Cont.) {.fragile}

```{r print_random_draw, echo=FALSE, tidy=TRUE}
message('Hand: \n', paste(random_draw, collapse = '\n'))
```

### Generating Distributions

-   What if we wanted to move beyond random selection where each draw or iteration exists with equal probability or within a uniform distribution?

    \par \vspace{5mm}

-   \texttt{R} is very flexible and capable of illustrating sampling distributions against expected outcomes

### Generating Distributions (Standard Normal)

-   Let's start with 1000 samples from a standard normal distribution where $\mu$ = 50 and $\sigma$ = 10

```{r normal_distribution, echo=TRUE, tidy=TRUE}
normal <- rnorm(1000, mean = 50, sd = 10) 
```

```{r normal_distribution_figure, echo=FALSE, tidy=TRUE, fig.height=2, fig.width=3}

data.frame(value = normal) %>%
  ggplot(aes(x = value)) + 
  geom_histogram(aes(y = after_stat(count)), fill = 'skyblue4', colour = 'black', bins = 30, alpha = 1/2) + 
  geom_vline(xintercept = 50, linetype = 2, colour = 'black', linewidth = 1) + 
  stat_function(fun = function(x) dnorm(x, mean = 50, sd = 10) * length(normal) * diff(range(normal))/30, 
                color = "red", size = 1) + 
  geom_hline(yintercept = 0) + 
  labs(x = ' ', y = ' ') + 
  scale_y_continuous(breaks = seq(25, 100, 25)) + 
  default_ggplot_theme + 
  theme(axis.text.y = element_blank())

```

### Generating Distributions (Standard Normal)

-   \textbf{Your Turn}: Generate 1000 draws from a standard normal distribution where $\mu$ = 25 and $\sigma$ = 10.

### Generating Distributions (Standard Normal -- Ex)

```{r normal_distribution_ex, echo=TRUE, tidy=TRUE}
normal <- rnorm(1000, mean = 25, sd = 10) 
```

```{r normal_distribution_ex_figure, echo=FALSE, tidy=TRUE, fig.height=2.25, fig.width=3.25}

data.frame(value = normal) %>%
  ggplot(aes(x = value)) + 
  geom_histogram(aes(y = after_stat(count)), fill = 'skyblue4', colour = 'black', bins = 30, alpha = 1/2) + 
  geom_vline(xintercept = 25, linetype = 2, colour = 'black', linewidth = 1) + 
  stat_function(fun = function(x) dnorm(x, mean = 25, sd = 10) * length(normal) * diff(range(normal))/30, 
                color = "red", size = 1) + 
  geom_hline(yintercept = 0) + 
  labs(x = ' ', y = ' ') + 
  scale_y_continuous(breaks = seq(25, 100, 25)) + 
  scale_x_continuous(breaks = seq(-5, 55, 10)) + 
  default_ggplot_theme + 
  theme(axis.text.y = element_blank())

```

### Generating Distributions (Exponential -- Ex)

-   \textbf{Your Turn}: Generate 1000 draws from an exponential distribution where \texttt{rate} = 2

### Generating Distributions (Exponential -- Ex)

```{r rexp_ex, echo=TRUE, tidy=TRUE}
exp <- rexp(1000, rate = 2)
```

```{r rexp_ex_figure, echo=FALSE, tidy=TRUE, fig.height=2.25, fig.width=3.25}

data.frame(value = exp) %>%
  ggplot(aes(x = value)) + 
  geom_histogram(aes(y = after_stat(density)), fill = 'skyblue4', colour = 'black', bins = 30, alpha = 1/2) +
  stat_function(fun = function(x) dexp(x, rate = 2), 
                color = "red", size = 1) + 
  geom_hline(yintercept = 0) + 
  labs(x = ' ', y = ' ') + 
  default_ggplot_theme + 
  theme(axis.text.y = element_blank())

```

# Functions

### Functions (Basics)

-   Functions are reusable blocks of code that perform a specific task when called, helping avoid repetition.

    \par \vspace{2.5mm}

-   They can take arguments (inputs) and return values (outputs), making them flexible and generalizable.

    \par \vspace{2.5mm}

-   They can also be combined, nested, and used within other functions to build complex workflows in a clear, organized way.

### Functions (Basic Syntax)

\small

```{r function_basic_syntax, eval=F, echo=T, warning=FALSE}

function_name <- function(input_1, input_2){
  
  #Code to Assume Input_1 and Input_2 
  
  #return(Return Output Value or Object)
  
}


```

### Functions (Example)

\small

```{r function_basic_example, eval=F, echo=T, warning=FALSE}

add_numbers <- function(x, y) {
  result <- x + y   
  return(result)   
}

add_numbers(5, 3)


```

### Functions (Basics, Cont.)

-   Take some time to try your own!
-   Try:
    -   Random Number Generation
    -   Easy Task Completion (e.g., addition, subtraction, etc.)

# Loops

### Loops (Basics)

-   In \texttt{R}, a for loop is a control structure used to repeat a block of code a fixed number of times, iterating over a sequence of values. The basic syntax is:

```{r basic_lor_loop_syntax, eval=FALSE, echo=TRUE}

for (variable in sequence){
  # Repeating Code Routine
  # return(Result Value or Object)
}

```

### Loops (Basics, Cont.)

-   For example, we can complete basics rolls of six-sided dice:

\footnotesize

```{r basic_for_loop_dice, tidy = T}

rolls <- c()

for (i in 1:10){
  
  temp_roll <- sample(1:6, 1, 
                      replace = TRUE, 
                      prob = rep(1/6, 6))
  
  rolls <- c(rolls, temp_roll)
}

rolls # Print

```

### Loops (Basics, Cont.)

-   We can also conditionally iterate through different values from the \texttt{functions} example

\footnotesize

```{r loop_function_example,  warning=FALSE}

add_numbers <- function(x, y) {
  result <- x + y   
  return(result)   
}

available_values <- c(1:10)
sums <- c()

for (pair in seq(1:10)){
  temp_pair <- sample(available_values, 2)
  sums <- c(sums, add_numbers(temp_pair[1], temp_pair[2]))
}

sums

```

### Loops (Basics, Cont.)

-   I can also deal 5 hands from a standard 52-card deck for a game of Texas Hold 'Em

    \par \vspace{2.5mm}

-   Here's the setup -- What's Next?

\footnotesize

```{r basic_for_loop_holdem, tidy = T}

set.seed(1234) # Seed
cards <- as.character(c(2:10, 'J', 'Q', 'K', 'A')) 
suits <- c('Hearts', 'Diamonds', 'Spades', 'Clubs') 
deck <- expand.grid(value = cards, 
                               suit = suits) |>
  mutate(card = paste(value, "of", suit)) |>
  pull(card) # Create a Full Deck

hands <- lapply(1:5, function(x) x) 

```

### Loops (Basics, Cont.)

\footnotesize

```{r basic_for_loop_holdem_cont, tidy = T}

for (card in 1:2){
  for (player in 1:5){
    temp_player_card <- sample(deck, 1, replace = F) 
    deck <- deck[!deck %in% temp_player_card]
    hands[[player]][card] <- temp_player_card
  } # For All 5 Players
} # For Both Cards

do.call(cbind, hands)

```

### Games of Chance: Blackjack

\centering

\textbf{What are the basic rules of Blackjack?}

\par \vspace{5mm}

\includegraphics[width=9cm]{../../images/blackjack.png}

### Rules of Blackjack:

-   Objective: Beat the dealer by getting closer to 21 without going over
-   Card values:
    -   Number Cards = Face Value
    -   Face Cards = 10 (Aces = 1 *or* 11)
-   Dealer Rules: Dealer reveals cards after players act and must hit until *at least* 17
-   Gameplay:
    -   Go Over 21 = **BUST** (Loss)

    -   Tie w/ Dealer = Push (No Win/Loss)

    -   Standard Win = **1:1** (Win Bet x2)

    -   Blackjack (Ace + 10-Value Card = **3:2**

### Blackjack Exercise

```{=tex}
\begin{center}
\textbf{Write an \texttt{R} routine to play a round of Blackjack. I will do the same.}
\end{center} \par \vspace{5mm}
```
-   *Hint*: Sample from all 52 cards without replacement...

### Blackjack Exercise (Cont.)

1.  What if we play with a four-deck shoe? \pause

    \par \vspace{2.5mm}

2.  What if I wanted to repeat this process 1,000 times?

    \par \vspace{2.5mm}

*Hint*: Use a loop!

### Blackjack Exercise (Cont.)

1.  Assume I begin with \$1000 every day and bet \$100 each game (though I'll only play 10 hands...). Over 100 days, approximately how much money am I left with? *Note*: If I run out of money on a given day, I'm done -- also, each day restarts with \$1000 but previous day's leftover sum is added to aggregate winnings.

    \par \vspace{2.5mm}

2.  What if I start with \$1000 but don't replace the money every day... How much will I have after 10 days? 50 days?

    \par \vspace{2.5mm}

3.  Take some time then play around with blackjack_simulation.R

### Roulette Exercise

-   Head over to Course GitHub (Intermediate Programming in \texttt{R})
-   Bottom of \texttt{Number Generation \& Loops}

# Data Visualization

-   \texttt{ggplot()} is an incredibly flexible visualization tool

    \par \vspace{2.5mm}

-   There's a balance between professional & \textit{noisy} visualizations

    \par \vspace{2.5mm}

-   Some journals & reviewers are more critical than others

    \par \vspace{2.5mm}

-   \textbf{Goal}: \textit{Publication-ready} visualizations capable of relaying inferential value on its own

### My Default ggplot() Aesthetics

\scriptsize

```{r default_ggplot_theme, eval=T, echo=T}

default_ggplot_theme  <- theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12),
    
    axis.title = element_text(size = 12, colour = 'black'),
    
    axis.text = element_text(size = 10, colour = 'black'),
    
    panel.background = element_rect(linewidth = 1, colour = 'black', fill = NA), 
    
    legend.position = 'bottom', 
    
    legend.background = element_rect(linewidth = 1, colour = 'black', fill = NA), 
  )

```

### My Default ggplot() Aesthetics

\scriptsize

```{r default_ggplot_application_1}

set.seed(1234)

sample_data <- tibble(x = c(1:100), 
                      y = rnorm(100, mean = 0.75, sd = 0.33))

summary(sample_data)

```

### My Default ggplot() Aesthetics

\scriptsize

```{r default_ggplot_application_2, fig.height=2.5, fig.width=3.5, tidy=TRUE}

sample_data %>%
  ggplot(aes(x = x, y = y)) + 
  geom_point() + 
  geom_smooth(method = 'lm', formula = 'y~x')

```

### Adding Default Aes

\scriptsize

```{r default_ggplot_application_3, echo=FALSE, fig.height=3, fig.width=4, tidy=TRUE}

sample_data %>%
  ggplot(aes(x = x, y = y)) + 
  geom_point(size = 0.75, alpha = 2/3) + 
  geom_smooth(method = 'lm', formula = 'y~x') + 
  default_ggplot_theme

```

# Stargazer

### Stargazer

\scriptsize

```{r stargazer, message=FALSE, warning=FALSE, tidy=TRUE, eval = F}

library(stargazer)
temp_lm <- lm(Sepal.Length ~ Sepal.Width + Petal.Length + Petal.Width, data = iris)

stargazer::stargazer(temp_lm, type = 'text', omit.stat = c("ser", 'f', 'adj.rsq'),  dep.var.caption  = "")


```

### Stargazer -- Text Example

\scriptsize

```{r stargazer_2, echo=FALSE, message=FALSE, warning=FALSE, tidy=TRUE}

temp_lm <- lm(Sepal.Length ~ Sepal.Width + Petal.Length + Petal.Width, data = iris)

stargazer::stargazer(temp_lm, type = 'text', omit.stat = c("ser", 'f', 'adj.rsq'),  dep.var.caption  = "")


```

### Stargazer -- Latex Example (type = 'latex')

\tiny
```{r stargazer_3, echo=FALSE, message=FALSE, warning=FALSE, tidy=TRUE}

temp_lm <- lm(Sepal.Length ~ Sepal.Width + Petal.Length + Petal.Width, data = iris)

stargazer::stargazer(temp_lm, type = 'latex', omit.stat = c("ser", 'f', 'adj.rsq'),  dep.var.caption  = "")


```

### Stargazer -- Latex Example Rendered

```{=tex}
\tiny
\begin{table}[!htbp] \centering 
  \caption{Rendered \texttt{stargazer} Table} 
  \label{} 
\begin{tabular}{@{\extracolsep{5pt}}lc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
\\[-1.8ex] & Sepal.Length \\ 
\hline \\[-1.8ex] 
 Sepal.Width & 0.651$^{***}$ \\ 
  & (0.067) \\ 
  & \\ 
 Petal.Length & 0.709$^{***}$ \\ 
  & (0.057) \\ 
  & \\ 
 Petal.Width & $-$0.556$^{***}$ \\ 
  & (0.128) \\ 
  & \\ 
 Constant & 1.856$^{***}$ \\ 
  & (0.251) \\ 
  & \\ 
\hline \\[-1.8ex] 
Observations & 150 \\ 
R$^{2}$ & 0.859 \\ 
\hline 
\hline \\[-1.8ex] 
\textit{Note:}  & \multicolumn{1}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
\end{tabular} 
\end{table}
```
### Stargazer -- Multiple Models

```{r stargazer_multiple_models, message=FALSE, warning=FALSE, include=FALSE, tidy=TRUE}

temp_lm <- lm(Sepal.Length ~ Sepal.Width + Petal.Length + Petal.Width, data = iris)
temp_lm2 <- lm(Sepal.Length ~ Sepal.Width:Petal.Length + Petal.Width, data = iris)

stargazer::stargazer(temp_lm, temp_lm2, type = 'latex', omit.stat = c("ser", 'f', 'adj.rsq'),  dep.var.caption  = "")


```

```{=tex}
\tiny
\begin{table}[!htbp] \centering 
  \caption{Rendered \texttt{stargazer} Table with 2 Models} 
  \label{} 
\begin{tabular}{@{\extracolsep{5pt}}lcc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
\\[-1.8ex] & \multicolumn{2}{c}{Sepal.Length} \\ 
\\[-1.8ex] & (1) & (2)\\ 
\hline \\[-1.8ex] 
 Sepal.Width & 0.651$^{***}$ &  \\ 
  & (0.067) &  \\ 
  & & \\ 
 Petal.Length & 0.709$^{***}$ &  \\ 
  & (0.057) &  \\ 
  & & \\ 
 Petal.Width & $-$0.556$^{***}$ & $-$0.311$^{***}$ \\ 
  & (0.128) & (0.114) \\ 
  & & \\ 
 Sepal.Width:Petal.Length &  & 0.185$^{***}$ \\ 
  &  & (0.017) \\ 
  & & \\ 
 Constant & 1.856$^{***}$ & 4.150$^{***}$ \\ 
  & (0.251) & (0.078) \\ 
  & & \\ 
\hline \\[-1.8ex] 
Observations & 150 & 150 \\ 
R$^{2}$ & 0.859 & 0.821 \\ 
\hline 
\hline \\[-1.8ex] 
\textit{Note:}  & \multicolumn{2}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
\end{tabular} 
\end{table}
```
### Stargazer -- Summary Data

\footnotesize

```{r stargazer_summary_data,  tidy=TRUE}

stargazer(iris, 
          type = "text", 
          summary = TRUE,
          title = "Summary of Iris Dataset")


```

### Visualization Example

-   Using the \texttt{mtcars} dataset – \texttt{library(mtcars)} – complete the following: \par \vspace{2.5mm}
    1.   Using \texttt{mpg} as the dependent variable, compile two models using \texttt{cyl}, \texttt{disp}, \texttt{hp}, and \texttt{wt} – the second should have an interaction between \texttt{disp} and \texttt{wt}. \par \vspace{1.5mm}
    2.  Produce a table using \texttt{stargazer} of the resulting models. \par \vspace{1.5mm}
    3.  Use \texttt{ggplot} to illustrate the distribution of each of the variables listed in (1).
    
    
# Looking Forward

### Looking Forward

- Homework: Problem Set (Class 2) \par \vspace{2.5mm}
- Next Class: Parallel Computing 
