---
title: "Wordscores"
subtitle: "POS6933: Computational Social Science"
author: "Truscott (Spring 2026)"
output:
  html_document:
    self_contained: false
    layout: null
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)

library(ggplot2)
library(dplyr)
library(cowplot)
library(stargazer)
library(doParallel)
library(parallel)
library(snow)
library(tm)
library(quanteda)
library(quanteda.textmodels)
library(gutenbergr)
library(stringr)
library(word2vec)
library(Rtsne)
library(scales)
library(textdata)
library(reticulate)

default_ggplot_theme <- theme_minimal() + 
  theme(panel.border = element_rect(linewidth = 1, colour = 'black', fill = NA), 
        axis.text = element_text(size = 12, colour = 'black'), 
        axis.title = element_text(size = 14, colour = 'black'), 
        strip.background = element_rect(linewidth = 1, colour = 'black', fill = 'grey75'),
        strip.text = element_text(size = 12, colour = 'black'), 
        legend.position = 'none', 
        plot.caption = element_text(hjust = 0, size = 10)) 


library(stringr)
library(textstem)
library(tm)

reduce_complexity <- function(text) {
  text <- tolower(text)
  text <- gsub("[^a-z\\s]", " ", text)
  words <- unlist(str_split(text, "\\s+"))
  words <- words[words != "" & !words %in% tm::stopwords("english")]
  words <- textstem::lemmatize_words(words)
  words <- words[str_detect(words, "^[a-z]+$")]
  text <- str_squish(paste(words, collapse = " "))
  return(text)
}


```

------------------------------------------------------------------------

## Wordscores (Laver, Benoit, and Garry)

Although probably (maybe?) not the first methodology for scaling ideology from texts, <code>Wordscores</code> --  introduced by Laver, Benoit, and Garry (2003) -- was the first to gain significant traction among political scientists. In a roundabout way, the methodology is a Naive Bayes–like generative model whose output is converted into a continuous ideological score rather than a class prediction. In essence, it's effectively (sort of) Naive Bayes + Value Projection onto a Reference Scale + Rescaling. 

Recall that for a standard (multinomial) Naive Bayes classifier, we first estimate how likely each word is to appear in each class $p(w|k)$. Then, for a new document ($D_{i}$), we use those learned word probabilities to calculate $\pi_{ik}$ -- the probability that $D_{i}$ belongs to each possible class $p(k\mid D_{i})$. The result is a classification assigned to the document because it has the highest probability. 

What <code>Wordscores</code> does instead is that after estimating how likely it is that each word belongs to a reference document ($p(w\mid r)$) -- e.g., a pair of left and right-aligned documents, we then calculate the expected ideological position of each word: 

$$ S_{w} = \sum_{r}p(r\mid w)A_{r} $$

Where $A_{r}$ is the ideological score of the reference text $r$. Rather than a categorical classification, the result is now a continuous ideological location. New documents $v$ are finally rescaled to account for weighted averages of word scores -- such that a document's position is the average ideology of the words it uses (from the reference documents) and weighted by the word frequences: 

$$ S_{v} = \sum_{w}p(w\mid v)S_{w} $$

### Wordscores -- Example (Manual)

Let's imagine we have two reference texts $r$ -- where $A_{r}$ is Left Party (-1) versus Right Party (+1) to represent the polarity of their ideologies. The two reference documents produce these word frequences: 

| Word | Left Ref. | Right Ref. | TOTAL |
| :----: | :----: | :----: | :---: |
| welfare | 8 | 1 | 9 |
| tax | 2 | 6 | 8 |
| military | 1 | 7 | 8 |
| <b>TOTAL</b> | 11 | 14 | 25 | 

Similar to Naive Bayes, we're going to compute the word probabilities for each of the polarities in the reference documents (classes): 

<b>welfare (9)</b>

$$p(\text{left}\mid \text{welfare}) = \frac{8}{9} = 0.889 \quad p(\text{right}\mid \text{welfare}) = \frac{1}{9} = 0.111  $$
<b>tax (8)</b>
$$p(\text{left}\mid \text{tax}) = \frac{2}{8} = 0.25 \quad p(\text{right}\mid \text{tax}) = \frac{2}{8} = 0.75 $$
<b>military (8)</b> 
$$p(\text{left}\mid \text{military}) = \frac{1}{8} = 0.125  \quad p(\text{right}\mid \text{military}) = \frac{7}{8} = 0.875 $$

Now we can compute the word scores using $S_{w} = \sum_{r}p(r\mid w)A_{r}$: 

$$S_{\text{welfare}} = 0.889(-1_{Neg}) + 0.111(1_{Pos}) = -0.778 $$
$$S_{\text{tax}} = 0.25(-1_{Neg}) + 0.75(1_{Pos}) = 0.50 $$

$$S_{\text{military}} = 0.125(-1_{Neg}) + 0.875(1_{Pos}) = 0.75 $$

We now have all the prior probabilities we need to start computing document scores for new documents ($v$). Suppose a new document contains the following: <b>Welfare</b> (3), <b>Tax</b> (1), and </b>Military</b> (0). From this, we know that $p(\text{welfare}\mid v) = \frac{3}{4} = 0.75$, and $p(\text{tax}\mid v) = \frac{1}{4} = 0.25$. 

To compute the document score using $S_{v} = \sum_{w}p(w\mid v)S_{w}$, we just plug in our new values and known probabilities from the reference documents: 

$$ S_{v} = 0.75(-0.778) + 0.25(0.5) = (-0.458)$$ 
```{r visualizing_wordscores, echo=FALSE}

data.frame(x = c(-1, -0.458, 1)) %>%
  ggplot() + 
    geom_vline(xintercept = 0, linetype = 2, alpha = 1/3) +
  geom_segment(aes(x = -1, xend = 1, y = 0, yend = 0), size = 1, colour = 'black') + 
  geom_point(aes(x = -0.458, y = 0), size = 5, shape = 21, colour = 'blue', fill = 'black') +
  geom_point(aes(x = -1, y = 0), size = 5, shape = 21, colour = 'black', fill = 'blue') +
  geom_point(aes(x = 1, y = 0), size = 5, shape = 21, colour = 'black', fill = 'red') +
    geom_text(label = 'New Document\n(Moderately Left)', x = -0.458, y = 0.015) + 
  geom_text(label = 'Left Ref.', x = -0.95, y = 0.025) + 
  geom_text(label = 'Right Ref.', x = 0.95, y = 0.025) + 
  scale_y_continuous(expand = c(0, 0), limits = c(-0.1, 0.1)) +
  theme_minimal() +
  theme(
    panel.background = element_rect(size = 1, colour = 'black', fill = NA), 
    panel.grid = element_blank(),        
    axis.text.y = element_blank(), 
    axis.title = element_blank(), 
    axis.text.x = element_text(size = 12, colour = 'black'),
    axis.ticks.y = element_blank()     
  ) 
```

### Wordscores -- Example (R Manual)

```{r, wordscores_manual}

A_r <- c(left = -1, right = 1) # Reference Ideologies

reference <- data.frame(word = c('welfare', 'tax', 'military'), 
                        left = c(8, 2, 1), 
                        right = c(1, 6, 7)) # Reference Documents

reference <- reference %>%
  mutate(p_left = reference$left/(reference$left + reference$right), 
         p_right = reference$right/(reference$right + reference$left), 
         p_left = round(p_left, 3), 
         p_right = round(p_right, 3)) # Assign P(r|w)

reference$S_w <- reference$p_left * A_r["left"] + reference$p_right * A_r["right"] # Calculate Word Scores

stargazer::stargazer(tibble(reference), summary = F, type = 'text')

new_document <- c(welfare = 3, 
                  tax = 1, 
                  military = 0) # New Document

new_document_total <- sum(new_document) # New Document Total Words

p_w_v = new_document/new_document_total # Prob. of Word from New Document

sum(p_w_v * reference$S_w) # Calculate Document Scores

```
### Wordscores -- Example (Quanteda)


```{r, wordscores_quanteda}

library(quanteda) # Load Quanteda

texts <- c(left_ref  = paste(c(rep('welfare', 8), rep('tax', 2), 'military'), collapse = " "),
           right_ref = paste(c('welfare', rep('tax', 6), rep('military', 7)), collapse = " "),
           new_doc = paste(c(rep('welfare', 3), 'tax'), collapse = " "))

dfm <- quanteda::dfm(quanteda::tokens(quanteda::corpus(texts))) # Create DFM from Tokenized Corpus

dfm

reference_scores <- c(-1, 1, NA) # Set Reference Scores

wordscores_model <- textmodel_wordscores(dfm, y = reference_scores)  # Run Wordscores Model

predict(wordscores_model, rescaling = "lbg") # Predict (Lavar, Benoit, and Garry Rescaling Method)

# Results Going to be a Bit Different Than Manual -- Rescaling with Small Corpora Tend to Stretch Scores, Often Overshooting Original Reference Anchors. Adding More Documents Will Improve Stability of Reference Anchors 


```

### Wordscores -- Example (Quanteda - Lots of Data)

```{r wordscores_german_example, message=FALSE, warning=FALSE}

corp_ger <- get(load('../data/german_manifestos.rdata')) # Load German Manifestos Corpus

corp_ger

dfm_ger <- quanteda::dfm(quanteda::tokens(corp_ger, remove_punct = T)) %>% # DFM from Tokenized Corpus
  dfm_remove(pattern = stopwords("de"))  # Remove German Stopwords

wordscores_german <- textmodel_wordscores(dfm_ger, 
                                         y = corp_ger$ref_score, 
                                         smooth = 1) # Apply Wordscors

summary(wordscores_german)

pred_ws <- predict(wordscores_german, se.fit = TRUE, newdata = dfm_ger) # Predict Scores from Unseen

quanteda.textplots::textplot_scale1d(pred_ws) # Quanteda Built-In Function


```
