---
title: "Class 8 Problem Set"
subtitle: "POS6933: Computational Social Science"
author: "Truscott (Spring 2026)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = TRUE,
	warning = TRUE
)
library(ggplot2)
library(dplyr)
library(cowplot)
library(stargazer)
library(doParallel)
library(parallel)
library(snow)
library(tm)
library(quanteda)
library(gutenbergr)
library(stringr)
library(rvest)

reduce_complexity <- function(text){
  text <- tolower(text) # Lower Case
  text <- tm::removePunctuation(text) # Punctuation
  text <- tm::removeNumbers(text) # Numbers
  text <- tm::removeWords(text, tm::stopwords("english")) # Stop Words
  text <- unlist(stringr::str_split(text, '\\s+')) # Tokenize 
  text <- textstem::lemmatize_words(text) # Lemmatize
  text <- paste(text, collapse = ' ') # Re-Append
  text <- gsub("\\s{2,}", ' ', text) # 2 or More Spaces --> One Space
  text <- trimws(text) # White Space
  return(text)
} # Function to Process Text for Bag of Word

library(topicmodels)


```

------------------------------------------------------------------------

<b>Note</b>: Students should always aim to produce <i>publication-worthy</i> tables and figures. Unless otherwise stated, tables should be rendered using <code>stargazer::()</code>, while figures can be rendered using <code>ggplot2::()</code> or <code>plot()</code>. Regardless, tables and figures should always be presented with necessary formatting -- e.g., (sub)title, axis (variable) labels and titles, a clearly-identifiable legend and key, etc. Problem sets must always be compiled using <code>LaTex</code> or <code>RMarkdown</code> and include the full coding routine (with notes explaining your implementation) used to complete each problem (10pts).

------------------------------------------------------------------------

1.  The sample data below represent terms recovered from American party platforms, where each row represents a party and terms from similar policy areas (4pt). 

::: {style="text-align:center;"}
| Document_ID | Category   | Terms                             |
| ----------- | ---------- | --------------------------------- |
| 1           | Democrat   | healthcare, taxes, security       |
| 2           | Democrat   | climate, energy, security         |
| 3           | Democrat   | healthcare, equity, border        |
| 4           | Democrat   | immigration, pathway, energy      |
| 5           | Democrat   | taxes, diplomacy, defense         |
| 6           | Republican | taxes, energy, security           |
| 7           | Republican | border, enforcement, energy       |
| 8           | Republican | healthcare, market, defense       |
| 9           | Republican | taxes, domestic, security         |
| 10          | Republican | immigration, enforcement, defense |

:::

  A. For two words found in both classes (<code>Democratic</code> and <code>Republican</code>), calculate the mutual information. There are <b>six</b> total, but you only need to choose <b>two</b>.
  
  B. For the words <code>Healthcare</code>, <code>Energy</code>, and <code>Security</code>, use the Fightin' Words approach to recover the log odds ratio that each is a predictor of either class. 
  
---
  
2. Using the <i>Federalist Papers</i> authored by <b>James Madison</b> (no co-authors or disputed authors), apply TF-IDF weighting and run k-means clustering using 5, 10, 15, 20, 25 cluster centers. Report the within-cluster sum of squares for each. What is the average reduction in the sum of squares for the addition of each cluster center? Is the answer the same if I ask this question for bins 5-15 versus 15-25? (3pt)


```{r fed_papers_clusters, eval=FALSE, include=FALSE}

fed_papers <- read_html('https://www.gutenberg.org/cache/epub/18/pg18-images.html')
essays <- html_elements(fed_papers, '.chapter')
text <- html_text2(essays)
text <- tibble(text)
federalist <- text %>%
  filter(stringr::str_detect(text, 'slightly different version', negate = TRUE)) %>%
  mutate(author = text %>%
           str_extract('HAMILTON AND MADISON|HAMILTON OR MADISON|HAMILTON|MADISON|JAY') %>%
           str_to_title(),
         title = str_extract(text, 'No. [A-Z].*')) %>%
  filter(author == 'Madison') # Clean & Partition

corpus <- VCorpus(VectorSource(sapply(federalist$text, reduce_complexity)))
dtm <- DocumentTermMatrix(corpus, control = list(weighting = weightTfIdf)) # TF-IDF DTM
dtm <- removeSparseTerms(dtm, 0.95)  # (Optional) Reducing Further Sparsity
dtm_matrix <- as.matrix(dtm) # Convert DTM Matrix for K-Means
cluster_centers <- seq(5, 25, 5)
k_means_output <- data.frame()

for (cluster in 1:length(cluster_centers)){
  temp_k_means <- kmeans(dtm_matrix, 
                         centers = cluster_centers[cluster], 
                         nstart = 25)
  k_means_output <- bind_rows(k_means_output, 
                              data.frame(cluster_center = cluster_centers[cluster], 
                                         sum_of_squares = temp_k_means$tot.withinss))
}


k_means_output$reduction <- c(NA, diff(k_means_output$sum_of_squares) * -1)
colnames(k_means_output) <- c("Cluster Center", "Sum of Squares", "Reduction")

stargazer::stargazer(k_means_output, 
                     summary=F, 
                     type = 'text')

mean(k_means_output$Reduction[2:3], na.rm = TRUE)  # 5 to 15
mean(k_means_output$Reduction[3:5], na.rm = TRUE)  # 15 to 25
```

3. Using the same <i>Federalist Papers</i> data (Madison authorship only), construct an LDA topic model with 10 topics. Print the top-10 words associated with each of those topics. 

  A. For each topic, report the posterior probabilities for the top 10 documents (words) in each. 
  
  B. For <b>at least 5</b> of the topics, write 2-3 sentences explaining what you think the "topic" identified by the model is -- i.e., what do these words tell you about the substantive content of the pooled words?  (3pt)
  
  


```{r, lda_topic, eval=FALSE, include=FALSE}

corpus <- VCorpus(VectorSource(sapply(federalist$text, reduce_complexity)))
dtm <- DocumentTermMatrix(corpus) # TF-IDF DTM

lda_federalist <- LDA(dtm, k = 10, method = "Gibbs", control = list(seed = 1234)) # LDA w/ 5 Topics & Gibbs Sampling 


topic_probs <- posterior(lda_federalist)$topics
print(round(topic_probs[c(1:10), c(1:5)], 2))

topicmodels::terms(lda_federalist, 10) # 10 Terms


```
