---
title: "Text Models for Measuring Ideology"
author: "Jake S. Truscott, Ph.D"
subtitle: |
  POS6933: Computational Social Science
institute: |
  \vspace{-5mm} University of Florida \newline
  Spring 2026 \newline \newline \newline
  \includegraphics[width=3cm]{../../../beamer_style/UF.png} \quad 
  \includegraphics[width=3.1cm]{../../../images/CSS_POLS_UF_Logo.png}
output: 
  beamer_presentation:
    slide_level: 3
    keep_tex: true
    toc: false
classoption: aspectratio=169  
header-includes:
  - \PassOptionsToClass{aspectratio=169}{beamer}
  - \usepackage{../../../beamer_style/beamer_style}
  - \setbeamersize{text margin left=3.5mm,text margin right=3.5mm}
---

```{r setup, include=FALSE}
library(dplyr); library(ggplot2); library(cowplot); library(formatR); library(stargazer); library(parallel); library(doParallel); library(tm); library(stringr); library(quanteda); library(quanteda.textstats); library(quanteda.textstats); library(dplyr); library(ggplot2); library(stringr);  library(tm); library(quanteda); library(quanteda.textstats); library(quanteda.textstats); library(rvest); library(tibble); library(tidytext); library(Rtsne)

knitr::opts_chunk$set(
  warning = FALSE, 
	fig.align = "center",
	comment = NA,
	dev = "pdf",
  tidy.opts=list(width.cutoff=50), 
  R.options = list(width = 120)
)

default_ggplot_theme  <- theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12),
    axis.title = element_text(size = 12, colour = 'black'),
    axis.text = element_text(size = 10, colour = 'black'),
    panel.background = element_rect(linewidth = 1, colour = 'black', fill = NA), 
    legend.position = 'bottom', 
    legend.title = element_blank(), 
    legend.background = element_rect(linewidth = 1, colour = 'black', fill = NA), 
  )

reduce_complexity <- function(text) {
  text <- tolower(text)
  text <- gsub("[^a-z\\s]", " ", text)
  words <- unlist(str_split(text, "\\s+"))
  words <- words[words != "" & !words %in% tm::stopwords("english")]
  words <- textstem::lemmatize_words(words)
  words <- words[str_detect(words, "^[a-z]+$")]
  text <- str_squish(paste(words, collapse = " "))
  return(text)
}


```

# Overview

### Overview
\begin{itemize}
\item Housekeeping
\item Class 9 Problem Set Review
\item \texttt{Wordscores}
\item \texttt{Wordfish}
\item \texttt{Wordshoal}
\item Office Hours (Until 6:00)
\end{itemize}

# Housekeeping

### Housekeeping
\begin{itemize}
\item \textbf{Reminders}:
  \begin{itemize}
    \item Slides due to me and peer reviewers by 11:59pm on Sunday (Schedule \emph{Next}) \par \vspace{2.5mm}
    \item Final materials (Paper, Appendix, and Final Slides) due 4/24 (Canvas)  \par \vspace{2.5mm}
    \item Rubric \& help document located on Canvas (\textit{Files}) -- Be sure to follow formatting requirements...
  \end{itemize}
  \item I will stay until 6:00 today if you want to chat about your papers
\end{itemize}


### Housekeeping -- Presentation Order
\centering
\scriptsize
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{llll}
\hline
\textbf{Day} & \textbf{Student} & \textbf{Reviewer 1} & \textbf{Reviewer 2} \\
\hline
4/12 & Arora, Manya & Mamun & Revilla, Valeria \\
4/12 & Cui, Lixiao & Meneley, Cory & Sarver, Sarah \\
4/12 & Giron-Khalil, Jailyn & Mohammed, Suren & Sultana, Abida \\
4/12 & Hao, Tianshu & Oneil, Connor & Tierman, Hunter \\
4/12 & Islam, Md Sahidul & Revilla, Valeria & Trenchi, Alejandro \\
4/12 & Mamun & Sarver, Sarah & Wang, Qianxi \\
4/12 & Meneley, Cory & Sultana, Abida & Arora, Manya \\
4/12 & Mohammed, Suren & Tierman, Hunter & Cui, Lixiao \\
4/19 & Oneil, Connor & Trenchi, Alejandro & Giron-Khalil, Jailyn \\
4/19 & Revilla, Valeria & Wang, Qianxi & Hao, Tianshu \\
4/19 & Sarver, Sarah & Arora, Manya & Islam, Md Sahidul \\
4/19 & Sultana, Abida & Cui, Lixiao & Mamun \\
4/19 & Tierman, Hunter & Giron-Khalil, Jailyn & Meneley, Cory \\
4/19 & Trenchi, Alejandro & Hao, Tianshu & Mohammed, Suren \\
4/19 & Wang, Qianxi & Islam, Md Sahidul & Oneil, Connor \\
\hline
\end{tabular}


### Class 9 Problem Set Review
\begin{itemize}
\item Notes...
\end{itemize}


# Text Models for Ideology

### Scaling Ideology 
```{r fig.height=1, fig.width=2, ideology_ex, include=F}

df <- data.frame(p = seq(-2, 2, 1), 
                 names = c('Very\nLeft', 'Left', 'Moderate', 'Right', 'Very\nRight'))

df %>%
  mutate(names = factor(names, levels = c('Very\nLeft', 'Left', 'Moderate', 'Right', 'Very\nRight'))) %>%
 ggplot(aes(x = p, y = 0)) + 
  geom_vline(xintercept = 0, linetype = 2, alpha = 1/3) +
  geom_segment(aes(x = -2, xend = 2, y = 0, yend = 0), size = 1, colour = 'black') + 
  geom_text(aes(label = names, y = 0.015)) + 
  geom_point(aes(x = p, y = 0, fill = factor(names)), colour = 'black', size = 3, shape = 21) + 
  scale_y_continuous(expand = c(0, 0), limits = c(-0.1, 0.1)) +
  scale_fill_manual(values = c('deepskyblue', 'deepskyblue', 'white', 'coral3', 'coral3')) + 
  theme_minimal() +
  theme(
    panel.background = element_rect(size = 1, colour = 'black', fill = NA), 
    panel.grid = element_blank(),        
    axis.text.y = element_blank(),
    axis.title = element_blank(), 
    axis.text.x = element_text(size = 12, colour = 'black'),
    axis.ticks.y = element_blank(), 
    legend.position = 'none'
  ) 

```


### Scaling Ideology (Cont.)
\begin{itemize}
\item \textbf{How I Was Taught to Think of It}: Rather than using right-hand side of the equation to solve for the left, flip the intuition -- Use DV to estimate optimal IV(s). \par \vspace{2.5mm} \pause
\item Assumptions: 
  \begin{enumerate}
  \item \textbf{Stable Latent Preferences}: Assumes actors have stable, coherent, underlying ideological positions that guide their behavior. \par \vspace{2.5mm} \pause
  \item \textbf{Behavior Reflects Ideology}: Treats votes or language as meaningful signals of those latent preferences (not pure noise or strategy). \par \vspace{2.5mm} \pause
  \item \textbf{Common Dimensional Structure}: Assumes actors operate in a shared ideological space (often low-dimensional) that is comparable across observations. 
  \end{enumerate}
\end{itemize}

### Scaling Ideology (Goals)
\begin{itemize}
\item \textbf{Recover Latent Position(s)}: Using observable behaviors as signals (ex: votes on legislation), estimate an underlying ideological dimension that isn't directly observed but we know exists and can be represented given both directionality (e.g., \emph{Left-Right}) and magnitude (e.g., \emph{Very Left}-\emph{Very Right}). \par \vspace{2.5mm} \pause
\item \textbf{Place Actors (Groups) On a Spectrum}: Relative location of actors (groups) in a common space denotes cardinality (and proximity) of relationship in that dimension.
\end{itemize}

### Text Models for Ideology (Benefits)
\begin{itemize}
\item Binary votes sometimes insufficient to fully understand elite political behaviors \par \vspace{2.5mm}
\item Text leverages high dimensional structure -- more variance and perhaps better explanatory power.  \par \vspace{2.5mm}
\item Text captures nuance in reasoning, framing, and emphasis that votes alone cannot reveal \par \vspace{2.5mm}
\item Enables dynamic and context-sensitive measurement of ideology across issues, time, and audiences
\end{itemize}

### Models Today
\begin{itemize}
\item Going to focus on 3 models: \texttt{Wordscores, Wordfish}, and \texttt{Wordshoal} \par \vspace{2.5mm}
\item \textbf{Important}: View subsequent models as methodological improvements on preceding \par \vspace{2.5mm}
\end{itemize}


# Wordscores

### Wordscores (Laver, Benoit, and Garry)
\begin{itemize}
\item Developed by Laver, Benoit, and Garry (2003) -- Probably not(?) first text-based model of ideology, but certainly first to gain significant use in social sciences. \par \vspace{2.5mm}
\item It’s effectively (sort of) Naive Bayes + Value Projection onto a Reference Scale + Rescaling \par \vspace{2.5mm} \pause 
\item \textit{Recall...} for standard Naive Bayes, we: 
\begin{enumerate}
  \item Estimate how likely each word $w$ is to appear in each class $k$ -- $p(w\mid k)$ \pause
  \item For new document $D_{i}$, we use those learned word probabilities to calculate $\pi_{ik}$ (probability that $D_{i}$ belongs in each possible class -- $p(k|mid D_i)$ \pause
  \item Classification assigned to the document because it has the highest probability
\end{enumerate}
\end{itemize}


### Wordscores (Laver, Benoit, and Garry) -- Cont.
\begin{itemize}
  \item \textbf{What \texttt{Wordscores} does instead}: Estimate how likely it is that each word belongs to a reference document -- $p(w\mid r)$, where $A_{r}$ represents the ideological score of those reference documents. \par \vspace{2.5mm} \pause 
  \item $ S_{w} = \sum_{r}p(r\mid w)A_{r} $ --  the score of some word $S_{w}$ is the sum of the probabilities it belongs to each reference document times the score associated with those documents. \par \vspace{2.5mm} \pause
  \item New documents $v$ are finally rescaled to account for weighted averages of word scores -- such that a document's position is the average ideology of the words it uses (from the reference documents) and weighted by the word frequences: 
  \item[] $$ S_{v} = \sum_{w}p(w\mid v)S_{w} $$
\end{itemize}

### Wordscores Example
\begin{columns}[T]
    \begin{column}{0.4\textwidth}
    \centering
    \begin{tabular}{lccc}
    \toprule
    Word & Left Ref. & Right Ref. & TOTAL \\
    \midrule
    welfare & 8 & 1 & 9 \\
    tax & 2 & 6 & 8 \\
    military & 1 & 7 & 8 \\
    \midrule
    \textbf{TOTAL} & 11 & 14 & 25 \\
    \bottomrule
    \end{tabular}
    \end{column}
    \begin{column}{0.6\textwidth}
    \begin{itemize}
        \item 3 Words (\emph{welfare, tax, military}) \par \vspace{2.5mm}
        \item 2 reference documents: 
        \begin{itemize}
          \item[] $A_{Left} = (-1)$ 
          \item[] $A_{Right} = (+1)$
        \end{itemize}
    \end{itemize}
    \end{column}
\end{columns}

### Wordscores Example (Cont.)
\begin{columns}[T]
    \begin{column}{0.4\textwidth}
    \centering
    \footnotesize
    \begin{tabular}{lccc}
    \toprule
    Word & Left Ref. & Right Ref. & TOTAL \\
    \midrule
    welfare & 8 & 1 & 9 \\
    tax & 2 & 6 & 8 \\
    military & 1 & 7 & 8 \\
    \midrule
    \textbf{TOTAL} & 11 & 14 & 25 \\
    \bottomrule
    \end{tabular}
    \end{column}
    \begin{column}{0.6\textwidth}
    \normalsize
    \begin{itemize}
        \item Let's start with word probabilities: 
        \item[] $p(\text{left}\mid \text{welfare}) = \frac{8}{9} = 0.889$ 
        \item[] $p(\text{right}\mid \text{welfare}) = \frac{1}{9} = 0.111$ \pause 
        \item[] $p(\text{left}\mid \text{tax}) =$ \pause $\frac{2}{8} = 0.25$ 
        \item[] $p(\text{right}\mid \text{tax}) =$ \pause $\frac{2}{8} = 0.75$
        \item[] $p(\text{left}\mid \text{military}) =$ \pause $\frac{1}{8} = 0.125$
        \item[] $p(\text{right}\mid \text{military}) =$ \pause $\frac{7}{8} = 0.875 $
    \end{itemize}
    \end{column}
\end{columns}

### Wordscores Example (Cont.)
\begin{columns}[T]
    \footnotesize
    \begin{column}{0.4\textwidth}
    \centering
    \begin{tabular}{lcc}
    \toprule
    Word & $p(\text{left}\mid w)$ & $p(\text{right}\mid w)$ \\
    \midrule
    welfare & 0.889 & 0.111 \\
    tax & 0.25 & 0.75 \\
    military & 0.125 & 0.875 \\
    \bottomrule
    \end{tabular}
    \end{column}
    \begin{column}{0.6\textwidth}
    \normalsize
    \begin{itemize}
        \item Now we can compute the word scores using 
        $S_{w} = \sum_{r}p(r\mid w)A_{r}$: \par \vspace{5mm}
        \item[] $S_{\text{welfare}} = 0.889(-1_{Neg}) + 0.111(1_{Pos}) = -0.778$ \par \vspace{2.5mm}
        \item[] $S_{\text{tax}}$ \pause $= 0.25(-1_{Neg}) + 0.75(1_{Pos}) = 0.50$ \par \vspace{2.5mm}
        \item[] $S_{\text{military}}$ \pause $= 0.125(-1_{Neg}) + 0.875(1_{Pos}) = 0.75$ 
    \end{itemize}
    \end{column}
\end{columns}


### Wordscores Example (Cont.)
\begin{itemize}
\item We can now start computing document scores for new documents $S_{v} = \sum_{w}p(w\mid v)S_{w}$ \par \vspace{2.5mm}
\item  Suppose a new document ($S_{v}$) contains the following: \textbf{Welfare} (3), \textbf{Tax} (1), and  \textbf{Military} (0).\par \vspace{2.5mm} 
\item We know that $p(\text{welfare}\mid v) = \frac{3}{4} = 0.75$, and $p(\text{tax}\mid v) = \frac{1}{4} = 0.25$. \par \vspace{2.5mm}
\item Solve for $S_{v}$ \pause $ = 0.75(-0.778) + 0.25(0.5) = (-0.458)$
\end{itemize}

### Wordscores Example (Cont.)
```{r visualizing_wordscores, echo=FALSE, fig.height=2, fig.width=5}

data.frame(x = c(-1, -0.458, 1)) %>%
  ggplot() + 
    geom_vline(xintercept = 0, linetype = 2, alpha = 1/3) +
  geom_segment(aes(x = -1, xend = 1, y = 0, yend = 0), size = 1, colour = 'black') + 
  geom_point(aes(x = -0.458, y = 0), size = 5, shape = 21, colour = 'blue', fill = 'black') +
  geom_point(aes(x = -1, y = 0), size = 5, shape = 21, colour = 'black', fill = 'blue') +
  geom_point(aes(x = 1, y = 0), size = 5, shape = 21, colour = 'black', fill = 'red') +
    geom_text(label = 'New Document\n(Moderately Left)', x = -0.458, y = 0.05) + 
  geom_text(label = 'Left Ref.', x = -0.95, y = 0.025) + 
  geom_text(label = 'Right Ref.', x = 0.95, y = 0.025) + 
  scale_y_continuous(expand = c(0, 0), limits = c(-0.1, 0.1)) +
  theme_minimal() +
  theme(
    panel.background = element_rect(size = 1, colour = 'black', fill = NA), 
    panel.grid = element_blank(),        
    axis.text.y = element_blank(), 
    axis.title = element_blank(), 
    axis.text.x = element_text(size = 12, colour = 'black'),
    axis.ticks.y = element_blank()     
  ) 
```


### Wordscores Example (Cont.)

\begin{itemize}
\item Recall: $S_{v} = \sum_{w}p(w\mid v)S_{w}$ \par \vspace{2.5mm}
\item  Suppose another new document ($S_{v}$) contains the following: \textbf{Welfare} (2), \textbf{Tax} (5), and  \textbf{Military} (3).\par \vspace{2.5mm} 
\item $p(\text{Welfare}\mid v) = 0.20$, $p(\text{Tax}\mid v) = 0.5$, and $p(\text{Military}\mid v) = 0.33$
\item Solve for $S_{v}$ \pause $ = 0.20(-0.778) + 0.5(0.25) + 0.33(0.75) = 0.217$ 
\end{itemize}

### Wordscores Example (Cont.)
```{r visualizing_wordscores_2, echo=FALSE, fig.height=2, fig.width=5}

data.frame(x = c(-1, -0.458, 1)) %>%
  ggplot() + 
    geom_vline(xintercept = 0, linetype = 2, alpha = 1/3) +
  geom_segment(aes(x = -1, xend = 1, y = 0, yend = 0), size = 1, colour = 'black') + 
  geom_point(aes(x = 0.217, y = 0), size = 5, shape = 21, colour = 'red', fill = 'black') +
  geom_point(aes(x = -1, y = 0), size = 5, shape = 21, colour = 'black', fill = 'blue') +
  geom_point(aes(x = 1, y = 0), size = 5, shape = 21, colour = 'black', fill = 'red') +
    geom_text(label = 'New Document\n(Moderately Right)', x = 0.217, y = 0.05) + 
  geom_text(label = 'Left Ref.', x = -0.95, y = 0.025) + 
  geom_text(label = 'Right Ref.', x = 0.95, y = 0.025) + 
  scale_y_continuous(expand = c(0, 0), limits = c(-0.1, 0.1)) +
  theme_minimal() +
  theme(
    panel.background = element_rect(size = 1, colour = 'black', fill = NA), 
    panel.grid = element_blank(),        
    axis.text.y = element_blank(), 
    axis.title = element_blank(), 
    axis.text.x = element_text(size = 12, colour = 'black'),
    axis.ticks.y = element_blank()     
  ) 
```



# Wordfish 
### Wordfish (Slapin and Proksch)
\begin{itemize}
\item \texttt{Wordfish} is a poisson factor model that (like Wordscores) tries to scale the ideological positions of documents. \par \vspace{2.5mm}  \pause 
\item It does \textbf{not} require reference documents -- it is \textbf{unsupervised} \par \vspace{2.5mm} 
\item Uses maximum likelihood to infer both a document's latent position $\theta_i$ and word discrimination $\beta_j$ (how strongly a word separates documents) \par \vspace{2.5mm} 
\item[] 
\end{itemize}

### Wordfish Specification
\begin{itemize}
\item[] $$x_{ij} \sim \text{Poisson}(\lambda_{ij})$$
\item[]  $$\lambda_{ij} = \exp(\alpha_{j} + \psi_j + \beta_j\theta_i) $$
\item[] Where: 
\footnotesize
  \begin{itemize}
  \item $x_{ij}$ = the count of word $j$ in document $i$ 
\item $\alpha_{i}$ = Document-specific verbosity (value to control for length of document -- ensures $\theta_i$ reflects content differences, not just document size) 
\item $\phi_j$ = Word-specific baseline frequency 
\item $\beta_j$ = Word-specific discrimination 
\item $\theta_i$ = Latent position of document $i$ (\textbf{What we're most interested in}!)
  \end{itemize}
\end{itemize}

